---
title: "Sparse Clustering Overview"
author: "Paul Harmon"
date: "August 29, 2018"
output: 
  prettydoc::html_pretty:
    theme: tactile
    highlight: github
---

```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri('msutrans.png'), 
               alt = 'logo', 
               style = 'position:absolute; top:1px; left:0px; padding:2px;height:100px;')

               
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2);library(dplyr);library(plotly)
library(tibble)
```

# Introduction
This document is designed to cover some of the high points of sparse clustering and its implementation in R using the sparcl. 

**A Note:** Sparcl package was removed from CRAN on 7/20/2018 "as check problems were not corrected despite reminders."(See page at https://cran.r-project.org/web/packages/sparcl/index.html.) However, there is another version out there by Kondo, Salibian-Bareera and Zamar (2016) called RSKC. 

# How does Sparse Clustering Work?



# Iplementations in R: 

```{r, fig.align = 'center'}
#installs and libraries the RSKC package
#install.packages('RSKC')
library(RSKC)

#installs and libraries the Sparcl package (possibly deprecated)
#install the most recent version then use R-studio to install from archived file
#install.packages("C:/Users/r74t532/Downloads/sparcl_1.0.3.tar.gz", repos = NULL, type = "source")
#devtools::install_version('sparcl',version = '1.0.3')
library(sparcl) #works on my laptop but not OPA computer
```



#Datasets 

## NBA Players
```{r}
nba <- read.csv('data/na.csv', header = TRUE)

#pca-based
pc1 <- prcomp(nba[,-c(1,2,3)], scale = TRUE, center = TRUE)
summary(pc1)

#do some clustering
library(mclust)
modclust <- mclustBIC(pc1$x[,1:2])
mc <- Mclust(pc1$x[,1:2], x = modclust)

#build a dataframe
dat1 <- tibble(pc1$x[,1],pc1$x[,2],mc$classification); names(dat1) <- c('s1','s2','class')

#build a plot
p <- ggplot(dat1) + geom_point(aes(s1,s2,color = factor(class))) + ggtitle("NBA Rookies 2017") + theme_classic()

ggplotly(p)


##################
##Compare the Sparse Clustering Methods to this: 
library(RSKC)
spk <- RSKC(nba[,-c(1,2,3)], ncl = 3, alpha = 0, L1 = 1) # Sparse K-Means
#see documentation but alpha = 0 and l1 = 1 gives sparse K means
#to get "robust" sparse k-means, we need alpha >0 and L1 = 1

rspk <- RSKC(nba[,-c(1,2,3)], ncl = 3, alpha = .5, L1 = 1) # Robust Sparse K-Means

dat1$spk <- spk$labels
dat1$rspk <- rspk$labels

#gives the sparse k-means 
p2 <- ggplot(dat1) + geom_point(aes(s1,s2,color = factor(spk))) + ggtitle("NBA Rookies 2017") + theme_classic()
ggplotly(p2)


#gives the robust sparse k-means
p3 <- ggplot(dat1) + geom_point(aes(s1,s2,color = factor(rspk))) + ggtitle("NBA Rookies 2017") + theme_classic()
ggplotly(p3)

```



## Carnegie Classifications
In this case, we have $n = 335 institutions$ with 8 different characteristics. 
```{r}

```









